\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2018-10-19 document/XSGRB_sample_rev1.tex
Last revisions before resubmission

The \textit{Neil Gehrels Swift Observatory} ({\it Swift}) satellite
\citep{Gehrels2004, Gehrels2009}, which was launched in 2004, has made it
possible to harvest much of the great potential in using GRBs as probes of the
intergalactic medium, which was already hinted at by results from earlier
missions \citep[e.g.,][]{Paradijs2000, Ricker2004}. With the three on-board
instruments, the Burst Alert Telescope \citep[BAT;][]{Barthelmy2005}, the X-Ray
Telescope \citep[XRT;][]{Burrows2005}, and the UltraViolet and Optical Telescope
\citep[UVOT;][]{Roming2005}, \swift~is an ideal observatory for GRB hunting. A
crucial aspect of the success of the {\it Swift} mission has been the extensive
ground-based follow-up observations of the afterglows and of the host galaxies
of the GRBs, involving a large community of researchers. This fruitful
collaboration has been facilitated by the open data access policy of the {\it
	Swift} mission. The close collaboration between detection facilities and
electromagnetic  follow-up campaigns continue to be immensely rewarding, as
recently highlighted by the simultaneous detection of gravitational waves and
light from the neutron star merger in the shape of GW170817/GRB170817A/AT2017gfo
\citep{2017ApJ...848L..14G, 2017ApJ...848L..15S, 2017ApJ...848L..16S, 2017ApJ...848L..17C, 2017ApJ...848L..18N, 2017ApJ...848L..19C, 2017ApJ...848L..27T, 2017ApJ...850L...1L, 2017Natur.551...64A, 2017Natur.551...67P, 2017Natur.551...71T, 2017Natur.551...75S, 2017Sci...358.1583K, LIGOScientificCollaboration2017a, LIGOScientificCollaboration2017}.

@@

	\item GRB trigger by BAT onboard the \swift~satellite

@@

We remark, however, that while the selection of the afterglows has been carried
out in a homogeneous way, the resulting quality of the spectra is not uniform.
Individual bursts vary significantly in brightness, due to intrinsic
differences in luminosity, dust extinction, and time of observation. The covered
rest-frame wavelength range is also different for each case. Given that our
observations are targets-of-opportunity, this leads to a varying S/N, and in
particular the sensitivity limit to detecting absorption features varies greatly
from burst to burst and as a function of wavelength.

@@

\begin{enumerate}
	\item The GRB triggered \textit{Swift} onboard.
	\item The X-ray position must be available less than 1 hr after the GRB.
	%    \item X-shooter is the currently operating instrument at the telescope.
	% DM I removed the line above because we do send the trigger even if X-shooter is not mounted (though, the trigger is rejected, as explained in the text above)
	%    \item GRB promptly visible from Cerro Paranal.
	%    \item Error radius < 60\arcsec.
	%    \item No limit on declination.
	%    \item No constraint on Moon phase or distance.
	\item The elevation of the source in the sky is $> 22^\circ$ (both at trigger time and 15 min after).
	\item The Sun elevation from Paranal is $< -12^\circ$ (both at trigger time and 15 min after).
	\item Galactic $A_V \leq 1.0$~mag according to the maps of \citet{Schlegel1998}.
	\item The \swift-circulated tags: KNOWN\_SOURCE, COSMIC\_RAY, DEF\_NOT\_GRB are set to false (PROB\_NOT\_GRB can be true)\footnote{These tags are distributed as part of the GCN notices. Explanations for all tags are \href{https://gcn.gsfc.nasa.gov/sock_pkt_def_doc.html}{available online}.}.
\end{enumerate}

@@

Due to a mechanical failure, the atmospheric dispersion corrector (ADC) was
disabled from 1st of August 2012 until the end of this program. Only GRB~100728B
was affected by the failing ADC prior to disablement, resulting in a
lower-than-nominal throughput. To avoid chromatic slit losses due to atmospheric
dispersion, nearly all subsequent observations have been carried out at
parallactic angle. A consequence of this is that for all observations following
1st of August 2012, the centroid of the trace of the source changes position
across the spatial direction of the slit as a function of wavelength. This
effect has been modeled in the extraction procedure, as described in Sect.
\ref{extract}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Slit loss correction and flux calibration accuracy} \label{fluxes}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For the majority of the observations presented in this sample, the slit width of
the target observations is less than the seeing PSF. This means that a fraction
of the incoming light is lost outside the slit. As an example, for atmospheric
conditions delivering a seeing PSF with a FWHM of 0\farc9 observed with a
0\farc9 slit, only 76.1 per cent of the light will enter the slit. This slit loss
is not corrected by the flux calibration, as flux standard stars are always
observed with a 5\arcsec{} slit. As a consequence, almost all spectra will,
without slit loss correction, underestimate the flux by some fraction.
Additionally, slit losses are wavelength dependent due to the wavelength
dependence of seeing \citep{Boyd1978} and as the seeing changes across the slit
in wavelength, so does the slit loss. This artificially introduces a change in
continuum slope.

\begin{figure}[!t]
	\centerline{\includegraphics[width=\columnwidth]{figures/flux_accuracy}}
\caption{The left panel shows the acquisition camera magnitudes as a function
	of synthetic photometry derived from the spectra. The blue, dashed line
	represent equality between the two measures. The right panel shows a histogram
	of the differences between the two. The vertical lines in the histogram
	indicates the 16th, 50th, and 84th percentiles of the distribution ($\Delta
	\mathrm{mag} = 0.07 ^{+0.51} _{-0.45} $). This shows the accuracy of the flux
	calibration. The colouring of the data-points shows the photometric
	passband used by the acquisition camera, as indicated by the labels. The $R$- and $I$- band magnitudes have been calibrated against the USNO-B1 catalog,
	which has a poor photometric accuracy; this contributes significantly to the  observed scatter, and for this reason we indicate these data with a light color.} \label{fig:flux_acq}
\end{figure}

To correct for the slitlosses, we develop an algorithm that can, based on the
atmospheric conditions at time of observation and the wavelength dependence of
seeing, estimate the losses at each wavelength bin. We start with the DIMM
seeing \citep{Sarazin1990}, which measures the seeing FWHM at zenith. We correct
the zenith seeing for the airmass of the observational sightline, which gives us
the effective seeing, delivered outside the slit at 5000 \AA. We can then use
that the seeing scales as $\sim \lambda^{-1/5}$ to obtain a measure of the
seeing across the entire wavelength coverage of X-shooter. At each wavelength
bin of the spectra, we simulate a synthetic 2D seeing PSF based on a normalised,
2D Moffat profile which we integrate over the used slit width. This can be used to correct for the estimated slit losses at each wavelength bin.

To assess the accuracy of the flux-calibration, after the slit loss correction,
we compare a subset of afterglow spectra with photometry derived from the
acquisition camera. The acquisition camera takes a short image just before
the spectroscopic integration, which only introduces a minimal delay and allows us to secure photometry for the vast majority of events. As photometric flux standards are rarely acquired, we resort to sky surveys such as the USNO catalog (for the $R$ and $I$ bands; \citealt{Monet2003}) or SDSS and Pan-STARRS ($griz$; \citealt{Abolfathi2018,Chambers2016}), where available.

@@

The effect of telluric correction varies greatly depending on the wavelength
region. In the red and ner-infrared areas, it allows to identify features in a
credible manner. which would be otherwise hidden among or blended with
atmospheric features. The telluric correction allows us to reliably recover the
continuum level even in regions of moderately low atmospheric transmission (see,
e.g., the A- and B-bands in Fig.~\ref{fig:spectrum}). We urge however to use
caution when deriving fitting lines in regions significantly affected by
telluric absorption.

@@

First, numerous works, including this one, have found that GRB 
absorption systems have much higher column densities than randomly 
selected sight lines 
\citep{2001ApJ...549L.209G,Prochaska2007, Fynbo2009}. 
This fact itself shows that the bulk of GRB absorption lines are not due 
to random intervening systems, but are peculiar to the GRB host. One 
second line of evidence comes from the systematic detection of nebular 
emission lines at the same redshift as the absorption system. 
\citet{Kruhler2015} report a 94\% detection rate of emission 
lines at $z < 1$ (where sensitivity is not an issue, even in the 
presence of a bright afterglow), always in agreement with absorption. 
Besides, it is extremely rare to detect emission from QSO intervening 
systems (see e.g. \citealt{2010MNRAS.408.2128F} for a discussion), which 
again indicates that GRB absorption systems are native to their hosts. 
An even more direct determination comes from the detection, in several 
cases, of fine-structure and metastable lines (for example from Fe and 
Ni), which are exclusively detected in GRB environments (and have been 
directly connected to the UV field generated by the GRB itself; e.g. 
\citealt{2006ApJ...648...95P, Vreeswijk2007, DElia2009}). 
Finally, at $z \gtrsim 2$ (where Ly$\alpha$ enters the observed range) 
the onset of the Ly$\alpha$ forest also provides an \textit{upper} limit 
on the GRB redshift, which again is always consistent with the 
absorption line redshift. The whole set of these arguments indicates 
that in only very few cases, if any at all, is a misidentification of 
the redshift with the intervening system a concrete possibility.

@@

12 of the $N_{\mathrm{HI}}$ measurements for these spectra have previously been
presented in \citet{Cucchiara2015} (See Table \ref{tab:HI}). We provide new
measurements here for completeness. In the compilation of $N_{\mathrm{HI}}$
measurements towards GRBs in Tanvir et al. (submitted) there are 93 published
$N_{\mathrm{HI}}$ values, excluding the measurements provided here. We here
provide 30 new neutral hydrogen column density measurements -- an increase of
the number of optically derived hydrogen column densities of $\sim$ 33 per cent.
We show the two distributions in  Fig. \ref{fig:NH_dist}. We compare the median,
the 16th, and 84th percentiles of the two distributions. The sample presented
has a \nh~$= 21.8_{-0.8}^{+0.3}$ and the rest of the literature values has
\nh~$= 21.5_{-1.5}^{+0.4}$. We see that the two distributions have a large
degree of overlap due to the large width of the distributions, but we find a
slightly higher median value for the new sample presented here. A 2-sided KS
test gives a $p$-value of p = 0.006, meaning relatively strong evidence against the
null hypothesis that the two samples are drawn from the same underlying distribution.
Because the bursts that have measurements of the hydrogen column density are
selected solely based on our ability to infer a column, it is difficult to make
any strong conclusions about the population statistics in terms of gas content.

In Fig. \ref{fig:NH_dist}, we also show the column density distribution
for the 12081 quasar absorbers with \nh~$> 20$ from \citet{Noterdaeme2012b}. The
fact that GRBs are systematically located behind the highest \nh, previously
noted \citep[e.g.,][]{Prochaska2007, Fynbo2009}, is very clear in this
figure. The reason for this is that quasar sample sight-lines through galaxies
that are cross-section selected, whereas GRB sight-lines probe the dense,
star-forming regions in their hosts. When comparing the GRB DLAs and and QSO DLAs we correct for the fact that, on average, the GRB light only goes though half the DLA. In Fig. \ref{fig:NH_dist}, we have therefore multiplied the inferred $N_{\mathrm{HI}}$ values by 2 \citep{Prochaska2008}.

\begin{figure}[!t]
	\centering \includegraphics[width=\columnwidth]{figures/NH_dist.pdf}
\caption{Distributions of hydrogen column densities for absorbers found in
	quasar absorption lines, from \citet{Noterdaeme2012b} in blue. Overplotted in
	green is the kernel density estimate of absorbers in GRB sightlines. Values are
	taken from the compilation in Tanvir et al. (submitted), along with the new
	values presented in this sample. We also show in the red the values derived in
	this work. The marginal distributions for the three samples are also shown
	along the left side and on the top, where the different environments probed are
	clearly visible in the hydrogen column densities, as previously also noted in
	\citet{Fynbo2009}. To correct for geometric differences between GRB DLAs and QSO DLAs, we have multiplied the inferred GRB DLA $N_{\mathrm{HI}}$ by 2.} \label{fig:NH_dist}
\end{figure}

@@

Our sample serves the purpose to further characterize the environments of GRBs
that was also much advanced by the previous surveys based primarily on
lower-resolution spectroscopy. GRB afterglow sight-lines are unique in the
sense that only after observing more than 12000 damped Lyman-$\alpha$ absorbers
(DLAs) towards about 10$^5$ quasars, a handful systems with \nh~$ > 22$ have been
identified \citep[e.g., five in][]{Noterdaeme2012b}. Long GRB afterglow spectra,
by contrast, reveal such systems routinely \citep[][and this
work]{Jakobsson2006b, Fynbo2009, Cucchiara2015}. With afterglow spectroscopy
(throughout the electromagnetic spectrum from X-rays to the sub-mm) we are able
to characterize the properties of star-forming galaxies over cosmic history in
terms of redshifts, metallicities, molecular contents, ISM temperatures, UV-flux
densities, extinction curves, etc.  A number of independent papers have been
published or submitted for publication focusing on many of these specific issues
of our sample such as extinction curves (\citealt{Japelj2015}, Zafar et al.
submitted, see also \citealt{Fynbo2014, Heintz2017a}), emission lines from the
underlying host galaxies \citep{Kruhler2015}, the frequency of intervening
\mgii~ absorbers \citep{Christensen2017}, \citet{Arabsalmani2018} on the
metallicity-scaling relations, and escape of ionizing radiation (Tanvir et al.,
submitted). A number of additional companion papers are also planned,
investigating the detailed properties of the sample presented here, including
equivalent width distributions (de Ugarte Postigo et al., in preparation),
metallicities and kinematics (Th{\"o}ne et al., in preparation), high ionization
lines (Heintz et al., in submitted), molecular lines (Bolmer et al., in
preparation), fine-structure lines (Vreeswijk et al., in preparation), and
composite GRB afterglow spectrum (Selsing et al., in preparation).

@@

\begin{figure*}[!h]
	\centering
	\includegraphics[page=1, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption{Measurements of the hydrogen column densities for all bursts with a
		clear Lyman alpha absorption system. In solid black is shown the spectrum with
		black dotted giving the corresponding 1-$\sigma$ error. The black dashed line shows zero
		flux density. The solid red line is the absorption of column density equal to
		the value presented in Tab. \ref{tab:HI} with the 1-$\sigma$ interval shown
		with dashed lines.}
	\label{fig:HI1}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=2, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI2}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=3, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI3}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=4, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI4}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=5, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI5}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=6, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI6}
\end{figure*}
\clearpage

@@

The data presented here also formed the basis of GCN 10620 \citep{GCN10620} and
GCN 10631 \citep{GCN10631} and are published in \citet{DeUgartePostigo2011}. The
burst has been followed up in three epochs of observations, 0.4, 1.4, and 2.4
days after the burst, each lasting 4.8 ks. The unambiguous redshift of the host,
$z=0.624$, is found from the simultaneous detection of emission features
belonging to nebular lines, including \hi, \oii, \oiii, \neiii, \nii, [\sii],
[\siii], and \hei~as well as absorption features due to the presence of \znii,
\crii, \feii, \mnii, \mgii, \mgi, \tiii, and \caii, all at a consistent
redshift. Temporal evolution of the fine structure lines belonging to \feii$^*$
is found between the epochs.

@@

The spectra presented here also formed the basis of GCN 10684 \citep{GCN10684}
and are used in Skuladottir (2010), but not published elsewhere. Observations
started 4 hours after the \textit{Swift} trigger, totaling 2.4 ks. Absorption
features from \mgii~and \feii~in the afterglow continuum are detected at
$z=1.1755$.

@@

The data presented here also formed the basis of GCN 10971 \citep{GCN10971}. The
spectrum has not otherwise been published previously. The observations were
carried out in RRM starting 11 min after the GRB trigger. See section \ref{RRM},
for a description of the RRM scheme. Absorption lines from several ionic species
are detected in the afterglow continuum at a common redshift of $z = 1.288$.

@@

This is a Fermi burst with a LAT detection and thus outside the statistical
sample, but nonetheless followed up due to the extremely high peak energy
\citep{Axelsson2012}. Starting 28.7 hours after the burst trigger, 2.4 ks
spectroscopic observation reveals after heavy binning, a wide, faint trace down
to $\sim$ 580 nm, offset by 2.5\arcsec~relative to the centering of the slit.
No good redshift measurement can be inferred from this. We have adopted the
redshift from GCN 12193 \citep{GCN12193}.

@@

This spectrum is empty, but is included here for completeness. It is not
published anywhere. $4\times600$ s spectroscopic observation, starting 3.7
hours after the trigger does not reveal anything conclusive. The host
association is additionally ambiguous for this, likely extincted \citep[GCN
17767;][]{GCN17767}, GRB.

>>>>

The \textit{Neil Gehrels Swift Observatory} ({\it Swift}) satellite
\citep{Gehrels2004, Gehrels2009}, which was launched in 2004, has made it
possible to harvest much of the great potential in using GRBs as probes of the
intergalactic medium, which was already hinted at by results from earlier
missions \citep[e.g.,][]{Paradijs2000, Ricker2004}. With the three on-board
instruments, the Burst Alert Telescope \citep[BAT;][]{Barthelmy2005}, the X-Ray
Telescope \citep[XRT;][]{Burrows2005}, and the UltraViolet and Optical Telescope
\citep[UVOT;][]{Roming2005}, \swift~is an ideal observatory for GRB hunting. A
crucial aspect of the success of the {\it Swift} mission has been the extensive
ground-based follow-up observations of the afterglows and of the host galaxies
of the GRBs, involving a large community of researchers. This fruitful
collaboration has been facilitated by the open data access policy of the {\it
	Swift} mission. The close collaboration between detection facilities and
electromagnetic  follow-up campaigns continue to be immensely rewarding, as
recently highlighted by the simultaneous detection of gravitational waves and
light from the neutron star merger in the shape of GW170817/GRB170817A/AT2017gfo
\citep{2017ApJ...848L..14G, 2017ApJ...848L..15S, 2017ApJ...848L..16S,
	2017ApJ...848L..17C, 2017ApJ...848L..18N, 2017ApJ...848L..19C,
	2017ApJ...848L..27T, 2017ApJ...850L...1L, 2017Natur.551...64A,
	2017Natur.551...67P, 2017Natur.551...71T, 2017Natur.551...75S,
	2017Sci...358.1583K, 2017Sci...358.1556C, Burgess2017, LIGOScientificCollaboration2017a,
	LIGOScientificCollaboration2017}.

@@

	\item GRB trigger by BAT on-board the \swift~satellite

@@

We remark, however, that while the selection of the afterglows has been carried
out in a homogeneous way, the resulting quality of the spectra is not uniform.
Individual bursts vary significantly in brightness, due to intrinsic
differences in luminosity, dust extinction, time of observation, etc. The covered
rest-frame wavelength range is also different for each case. Given that our
observations are targets-of-opportunity, this leads to a varying S/N, and in
particular the sensitivity limit to detecting absorption features varies greatly
from burst to burst and as a function of wavelength.

@@

\begin{enumerate}
	\item The GRB triggered \textit{Swift}.
	\item The X-ray position must be available less than 1 hr after the GRB.
	%    \item X-shooter is the currently operating instrument at the telescope.
	% DM I removed the line above because we do send the trigger even if X-shooter is not mounted (though, the trigger is rejected, as explained in the text above)
	%    \item GRB promptly visible from Cerro Paranal.
	%    \item Error radius < 60\arcsec.
	%    \item No limit on declination.
	%    \item No constraint on Moon phase or distance.
	\item The elevation of the source in the sky is $> 22^\circ$ (both at trigger time and 15 min after).
	\item The Sun elevation from Paranal is $< -12^\circ$ (both at trigger time and 15 min after).
	\item Galactic $A_V \leq 1.0$~mag according to the maps of \citet{Schlegel1998}.
	\item The \swift-circulated tags: KNOWN\_SOURCE, COSMIC\_RAY, DEF\_NOT\_GRB are set to false (PROB\_NOT\_GRB can be true)\footnote{These tags are distributed as part of the GCN notices. Explanations for all tags are \href{https://gcn.gsfc.nasa.gov/sock_pkt_def_doc.html}{available online}.}.
\end{enumerate}

@@

Due to a mechanical failure, the atmospheric dispersion corrector (ADC) was
disabled from 1st of August 2012 until the end of this program. Only GRB~100728B
was affected by the failing ADC prior to disablement, resulting in a
lower-than-nominal throughput. To avoid chromatic slit-loss due to atmospheric
dispersion, nearly all subsequent observations have been carried out at
parallactic angle. A consequence of this is that for all observations following
1st of August 2012, the centroid of the trace of the source changes position
across the spatial direction of the slit as a function of wavelength. This
effect has been modeled in the extraction procedure, as described in Sect.
\ref{extract}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Slit-loss correction and flux calibration accuracy} \label{fluxes}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

For the majority of the observations presented in this sample, the slit width of
the target observations is less than the seeing PSF. This means that a fraction
of the incoming light is lost outside the slit. This slit-loss is not corrected
by the flux calibration, as flux standard stars are always observed with a
5\arcsec{} slit. As a consequence, almost all spectra will, without slit-loss
correction, underestimate the flux by some fraction. Additionally, slit-loss
is wavelength dependent due to the wavelength dependence of seeing
\citep{Boyd1978} and as the seeing changes across the slit in wavelength, so
does the slit-loss. This artificially introduces a change in continuum slope.

\begin{figure}[!t]
	\centerline{\includegraphics[width=\columnwidth]{figures/flux_accuracy}}
\caption{The left panel shows the acquisition camera magnitudes as a function
	of synthetic photometry derived from the spectra. The blue, dashed line
	represent equality between the two measures. The colouring of the data-points
	shows the photometric passband used by the acquisition camera, as indicated by
	the labels. The $R$- and $I$- band magnitudes have been calibrated against the
	USNO-B1 catalog, which has a poor photometric accuracy; this contributes
	significantly to the observed scatter, and for this reason we indicate these
	data with a light color.  The right panel shows a histogram of the differences
	between the two. The vertical lines in the histogram indicates the 16th, 50th,
	and 84th percentiles of the distribution ($\Delta \mathrm{mag} = 0.07 ^{+0.51}
	_{-0.45} $). This shows the accuracy of the flux calibration.}
\label{fig:flux_acq}
\end{figure}

To correct for the slit-loss, we develop an algorithm that can, based on the
atmospheric conditions at time of observation and the wavelength dependence of
seeing, estimate the losses at each wavelength bin. We start with the DIMM
seeing \citep{Sarazin1990}, which measures the seeing FWHM at zenith. We correct
the zenith seeing for the airmass of the observational sightline, which gives us
the effective seeing, delivered outside the slit at 5000 \AA. We can then use
that the seeing scales as $\sim \lambda^{-1/5}$ to obtain a measure of the
seeing across the entire wavelength coverage of X-shooter. At each wavelength
bin of the spectra, we simulate a synthetic 2D seeing PSF based on a normalised,
2D Moffat profile which we integrate over the used slit width. This can be used to correct for the estimated slit-loss at each wavelength bin.

To assess the accuracy of the flux-calibration, after the slit-loss correction,
we compare a subset of afterglow spectra with photometry derived from the
acquisition camera. The acquisition camera takes a short image just before the
spectroscopic integration, which only introduces a minimal delay and allows us
to secure photometry for the vast majority of events. As photometric flux
standards are rarely acquired, we resort to sky surveys such as the USNO catalog
(for the $R$ and $I$ bands; \citealt{Monet2003}), SDSS, or Pan-STARRS ($griz$;
\citealt{Abolfathi2018,Chambers2016}), where available. The acquisition camera
is equipped with similar filters, so we can avoid transformation between filter
curves when comparing catalogues.

@@

The effect of telluric correction varies greatly depending on the wavelength
region. In the red and near-infrared areas, it allows to identify features in a
credible manner. which would be otherwise hidden among or blended with
atmospheric features. The telluric correction allows us to reliably recover the
continuum level even in regions of moderately low atmospheric transmission (see,
e.g., the A- and B-bands in Fig.~\ref{fig:spectrum}). Nonetheless, we recommend
caution when fitting lines in regions significantly affected by telluric
absorption.

@@

First, numerous works, including this one, have found that GRB absorption
systems have much higher column densities than randomly selected sight lines
\citep{2001ApJ...549L.209G, Savaglio2003,  Prochaska2007, Fynbo2009}. This fact
itself shows that the bulk of GRB absorption lines are not due to random
intervening systems, but are peculiar to the GRB host. A second line of evidence
comes from the systematic detection of nebular emission lines at the same
redshift as the absorption system. \citet{Kruhler2015} report a 94\% detection
rate of emission lines at $z < 1$ (where sensitivity is not an issue, even in
the presence of a bright afterglow), always in agreement with absorption.
Besides, it is rare to detect emission from QSO intervening systems
(see e.g. \citealt{2010MNRAS.408.2128F} for a discussion), which again indicates
that GRB absorption systems are native to their hosts. An even more direct
determination comes from the detection, in several cases, of fine-structure and
metastable lines (for example from Fe and Ni), which are exclusively detected in
GRB environments (and have been directly connected to the UV field generated by
the GRB itself; e.g. \citealt{2006ApJ...648...95P, Vreeswijk2007, DElia2009}).
Finally, at $z \gtrsim 2$ (where Ly$\alpha$ enters the observed range) the onset
of the Ly$\alpha$ forest also provides an \textit{upper} limit on the GRB
redshift, which again is always consistent with the absorption line redshift.
The whole set of these arguments indicates that in only very few cases, if any
at all, is a misidentification of the redshift with the intervening system a
concrete possibility.

@@

12 of the $N_{\mathrm{HI}}$ measurements for these spectra have previously been
presented in \citet{Cucchiara2015} (See Table \ref{tab:HI}). We provide new
measurements here for completeness. In the compilation of $N_{\mathrm{HI}}$
measurements towards GRBs in \citet{Tanvir2017} there are 93 published
$N_{\mathrm{HI}}$ values, excluding the measurements provided here. We here
provide 30 new neutral hydrogen column density measurements -- an increase of
the number of optically derived hydrogen column densities of $\sim$ 33 per cent.
We show the two distributions in  Fig. \ref{fig:NH_dist}. We compare the median,
the 16th, and 84th percentiles of the two distributions. The sample presented
has a \nh~$= 21.8_{-0.8}^{+0.3}$ and the rest of the literature values has
\nh~$= 21.5_{-1.5}^{+0.4}$. We see that the two distributions have a large
degree of overlap due to the large width of the distributions, but we find a
slightly higher median value for the new sample presented here. A 2-sided KS
test gives a $p$-value of p = 0.006, meaning relatively strong evidence against the
null hypothesis that the two samples are drawn from the same underlying distribution.
Because the bursts that have measurements of the hydrogen column density are
selected solely based on our ability to infer a column, it is difficult to make
any strong conclusions about the population statistics in terms of gas content.

In Fig. \ref{fig:NH_dist}, we also show the column density distribution for the
12081 quasar absorbers with \nh~$> 20$ from \citet{Noterdaeme2012b}. The fact
that GRBs are systematically located behind the highest \nh, previously noted
\citep[e.g.,][]{Prochaska2007, Fynbo2009}, is very clear in this figure. The
reason for this is that quasar sample sight-lines through galaxies that are
cross-section selected, whereas GRB sight-lines probe the dense, star-forming
regions in their hosts. When comparing the GRB DLAs and QSO DLAs we correct for
the fact that, on average the GRB light only goes through half of the full ISM
gas column. In Fig. \ref{fig:NH_dist}, we have therefore multiplied the inferred
$N_{\mathrm{HI}}$ values by 2 \citep{Prochaska2008}.

\begin{figure}[!t]
	\centering \includegraphics[width=\columnwidth]{figures/NH_dist.pdf}
\caption{Distributions of hydrogen column densities for absorbers found in
	quasar absorption lines, from \citet{Noterdaeme2012b} in blue. Overplotted in
	green is the kernel density estimate of absorbers in GRB sightlines. Values are
	taken from the compilation in \citet{Tanvir2017}, along with the new
	values presented in this sample. We also show in the red the values derived in
	this work. The marginal distributions for the three samples are also shown
	along the left side and on the top, where the different environments probed are
	clearly visible in the hydrogen column densities, as previously also noted in
	\citet{Fynbo2009}. To correct for geometric differences between GRB DLAs and QSO DLAs, we have multiplied the inferred GRB DLA $N_{\mathrm{HI}}$ by 2.} \label{fig:NH_dist}
\end{figure}

@@

Our sample serves the purpose to further characterize the environments of GRBs
that was also much advanced by the previous surveys based primarily on
lower-resolution spectroscopy. GRB afterglow sight-lines are unique in the sense
that only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ quasars, a handful systems with \nh~$ > 22$ have been
identified \citep[e.g., five in][]{Noterdaeme2012b}. Long GRB afterglow spectra,
by contrast, reveal such systems routinely \citep[][and this
work]{Jakobsson2006b, Fynbo2009, Cucchiara2015}. With afterglow spectroscopy
(throughout the electromagnetic spectrum from X-rays to the sub-mm) we are able
to characterize the properties of star-forming galaxies over cosmic history in
terms of redshifts, metallicities, molecular contents, ISM temperatures, UV-flux
densities, extinction curves, etc.  A number of independent papers have been
published or submitted for publication focusing on many of these specific issues
of our sample such as extinction curves (\citealt{Japelj2015}, Zafar et al.
submitted, see also \citealt{Fynbo2014, Heintz2017a}), emission lines from the
underlying host galaxies \citep{Kruhler2015}, the frequency of intervening
\mgii~ absorbers \citep{Christensen2017}, \citet{Arabsalmani2018} on the
metallicity-scaling relations, and escape of ionizing radiation
\citep{Tanvir2017} . A number of additional companion papers are also planned,
investigating the detailed properties of the sample presented here, including
equivalent width distributions (de Ugarte Postigo et al., in preparation),
metallicities and kinematics (Th{\"o}ne et al., in preparation), high ionization
lines (Heintz et al., in submitted), molecular lines \citep{Bolmer2018},
fine-structure lines (Vreeswijk et al., in preparation), and composite GRB
afterglow spectrum (Selsing et al., in preparation).

@@

\begin{figure*}[!h]
	\centering
	\includegraphics[page=1, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption{Measurements of the hydrogen column densities for all bursts with a
		clear Lyman alpha absorption system. In solid black is shown the spectrum with
		black dotted giving the corresponding 1-$\sigma$ error. The black dashed line shows zero
		flux density. The solid red line is the absorption of column density equal to
		the value presented in Tab. \ref{tab:HI} with the 1-$\sigma$ interval shown
		with dashed lines. The pressence of \lya, as noted in Tab. \ref{tab:HI}, is indicated with a black arrow.}
	\label{fig:HI1}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=2, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI2}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=3, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI3}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=4, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI4}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=5, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI5}
\end{figure*}
\clearpage
\begin{figure*}[!h]
	\centering
	\includegraphics[page=6, width=0.85\linewidth]{figures/HI_measurements.pdf}
	\caption*{Fig. \ref{fig:HI1}. continued.}
	\label{fig:HI6}
\end{figure*}
\clearpage

@@

The data presented here also formed the basis of GCN 10620 \citep{GCN10620} and
GCN 10631 \citep{GCN10631} and are published in \citet{DeUgartePostigo2011}. The
burst has been followed up in three epochs of observations, 0.4, 1.4, and 2.4
days after the burst, each lasting 4.8 ks. The unambiguous redshift of the host,
$z=0.624$, is found from the simultaneous detection of emission features
belonging to nebular lines, including \hi, \oii, \oiii, \neiii, \nii, [\sii],
[\siii], and \hei~as well as absorption features due to the presence of \znii,
\crii, \feii, \mnii, \mgii, \mgi, \tiii, and \caii, all at a consistent
redshift. Temporal evolution of the fine-structure lines belonging to \feii$^*$
is found between the epochs.

@@

The spectra presented here also formed the basis of GCN 10684 \citep{GCN10684}
and are used in Skuladottir (2010), but not published elsewhere. Observations
started 4 hours after the \textit{Swift} trigger, totalling 2.4 ks. Absorption
features from \mgii~and \feii~in the afterglow continuum are detected at
$z=1.1755$.

@@

The data presented here also formed the basis of GCN 10971 \citep{GCN10971}. The
spectrum has not otherwise been published previously. The observations were
carried out in RRM starting 11 minutes after the GRB trigger. See section \ref{RRM},
for a description of the RRM scheme. Absorption lines from several ionic species
are detected in the afterglow continuum at a common redshift of $z = 1.288$.

@@

This is a Fermi burst with a LAT detection and thus outside the statistical
sample, but nonetheless followed up due to the extremely high peak energy
\citep{Axelsson2012}. Starting 28.7 hours after the burst trigger, 2.4 ks
spectroscopic observation reveals after heavy binning, a wide, faint trace down
to $\sim$ 580 nm, offset by 2.5\arcsec~relative to the centring of the slit. No
good redshift measurement can be inferred from this. We have adopted the
redshift from GCN 12193 \citep{GCN12193}.

@@

This spectrum is empty, but is included here for completeness. It is not
published anywhere. $4\times600$ s spectroscopic observation, starting 3.7
hours after the trigger does not reveal anything conclusive. The host
association is additionally ambiguous for this, likely reddened \citep[GCN
17767;][]{GCN17767}, GRB.
